---
title: "diffloop: Identifying differential DNA loops from ChIA-PET data"
author: "Caleb Lareau & Martin Aryee"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{diffloop: Identifying differential DNA loops from ChIA-PET data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(diffloop)
library(diffloopdata)
library(ggplot2)
library(GenomicRanges)
```

## Preprocessing
The raw FASTQ read files were preprocessed with the `dnaloop`
(https://github.com/aryeelab/dnaloop) package, a `PyPi` package that relies 
on `samtools`, `bedtools`, `cutadapt`, and `MACS2` to align reads,
call anchor peaks, and summarize PETs per sample in the ChIA-PET experiment.
<br> <br>
To use the `looptestMake` function from a different preprocessing step,
have files `X.loop_counts.bedpe`,`Y.loop_counts.bedpe`, 
`Z.loop_counts.bedpe` in `bed_dir` for `samples = (X,Y,Z)` where the 
first lines should resemble: <br> <br>
1 10002272 10004045 10 120968807 120969483 . 1 <br>
1 10002272 10004045 10 99551498 99552470 . 1 <br>
1 10002272 10004045 1 10002272 10004045 . 17 <br>
<br> <br>
where the first three columns specify the position (chr:start:stop) of
the first anchor, the second three columns specify the position
(chr:start:stop) of the secnd anchor, the 7th column is "." and the 8th
column is the number of paired-end reads support that particular PET.

## Loading data
We read in data from the preprocessing pipeline using the `loopdataMake`
function. The example below uses sample data included in the `diffloopdata`
package. It contains interactions involving chromosome 1 from a set of six
Cohesin ChIA-PET samples.$^{1,2}$ You would normally set `bed_dir` to your 
real data directory.

``` {r load_data} 
bed_dir <- system.file("extdata", "esc_jurkat", package="diffloopdata")
bed_dir
samples <- c("naive_esc_1", "naive_esc_2", "primed_esc_1", "primed_esc_2",
            "jurkat_1", "jurkat_2")
```
Here, we use 1500 bp as a value to merge nearby anchors into
a single peak. In the `loopdataMake` function, the default is zero and should
be parameterized based on the resolution of the ChIA-PET Experiment.
```{r read_data}
full <- loopdataMake(bed_dir, samples, 1500)
celltypes <- c("naive", "naive", "primed", "primed", "jurkat", "jurkat")
full <- updateLDGroups(full, celltypes)
head(full, 3)
```
Alternatively, one could run `full <- loopdataMake(bed_dir)` and the `samples`
would be inferred from the directory contents. The default call to `loopdataMake` 
does not merge nearby anchors but instead keeps them distinct. If we want to 
group nearby anchors after the initial data read, one can run the `mergeAnchors`
command with the appropriate gap value. <br><br>
We'll now look at a PCA plot of the raw data:
```{r pca_prenorm, fig.width=7, fig.height=4, fig.cap ="PCA plot of**un-normalized** ChIA-PET loop data"}
pcaPlot(full) + geom_text(aes(label=samples)) 
```
From this initial PCA plot, it's not clear how the samples separate based on
their topologies, suggesting some quality control may be warranted.

## QC
First, we normalize the `loopdata` object, which updates the
`sizeFactor` of each sample by the number of reads identical to the
normalization technique employed in `DESeq2`. We recommend computing
the size factors on the original `loopdata` object to generate the best
normalization factors. 

```{r sizeFactors}
full <- calcLDSizeFactors(full)
head(full,3)
```

Looking at the full dataset, we can see that a majority of the counts
were within self-loops:
```{r qc}
loopMetrics(full)
```
We define "valid" looping in a combined experiment as those supported
by at least 3 replicates in 2 or more samples. We can generate valid 
loops by using the `filterLoops` command 
```{r qc2}
full_5kb <- filterLoops(full, width=5000, nreplicates=3, nsamples = 1)
dim(full_5kb)
valid_full <- filterLoops(full, width=5000, nreplicates=3, nsamples = 2)
dim(valid_full)
```
We can also see the proportion
of loops that occur within the same chromosome (intrachromosomal) or 
have loops that bridge chromosomes (interchromosomal). 
```{r qc3}
dim(intrachromosomal(valid_full))
dim(interchromosomal(valid_full))
```
Finally, we can determine the number of anchors supported by each sample 
as well as the loops with different thresholds of counts.
```{r qc4}
numLoops(valid_full,1:5)
head(valid_full)
```

## Post-Normalization

```{r pca_postnorm, fig.width=7, fig.height=4, fig.cap = "PCA plot of normalized ChIA-PET loop data"}
pcaPlot(valid_full) + geom_text(aes(label=samples)) +
    expand_limits(x = c(-20, 34))
```
Here, the PC plot is much better as the first principal component separates
the stem cells and the jurkat samples while the second PC separates the two
stem cell samples.

## Differential Looping
Here, we describe the most general association method first fitting the model
then doing pairwise comparisons by our choice of parameterizing the loopTest
function.
```{r geneAssoc}
# Fit edgeR Models Pairwise between cell types
fit <- loopFit(valid_full)
jn_res <- loopTest(fit,coef=2)
jp_res <- loopTest(fit,coef=3)
np_res <- loopTest(fit,contrast=c(0,-1,1))

head(np_res)
```

We could also have computed pairwise associations by subsetting the data and
using quickAssoc, which only allows loopdata objects with two group classes.

```{r quickAssoc}
np <- valid_full[,1:4]
summarize(np[1:3,])
np_res2 <- quickAssoc(np)
```

Instead of determining the loops that distinguish these conditions from each
other pairwise, this framework allows us to determine the loops that are
significantly different between all of these conditions:

```{r assocResults}
jnp_res <- loopTest(fit,coef=2:3)
head(jnp_res)
#Can also directly subset diffloop objects
jnp_res[100:101,]
```


## Differential Looping Regions
What regions are significantly different between jurkat and naive? We can 
answer this by doing a sliding window test by specifying the window size
and the step size in the `slidingWindowTest` function.
```{r swAssoc}
sw_jn <- slidingWindowTest(jn_res,200000,200000)
head(sw_jn)
```

What gene bodies have significantly different looping between the cell lines?
Whereas the `slidingWindowTest` makes no prior assumptions about the regions
to be tested, `featureTest` requires a `GRanges` object of coordinates to 
determine what regions will be tested for association.
```{r featAssoc}
chr1 <- getHumanGenes(c("1"))
ft_jnp <- featureTest(jnp_res,chr1)
head(ft_jnp,10)
```

## Quantifying Associations
How many differential loops are present pairwise between these conditions?
```{r pairwiseSum, message=FALSE, warning=FALSE}
np_sig_res <- topLoops(np_res, FDR = 0.2)
dim(np_sig_res)
dim(topLoops(jp_res, FDR = 0.2))
dim(topLoops(jn_res, FDR = 0.2))
summarize(np_sig_res)
```
At FDR < 0.2, only 6 loops are significantly different from the naive and
primed stem cells whereas the jurkat differs between the primed stem cells
by 177 significant loops and the naive by 211 significant loops on chromosome 1.
These results are in line with the underlying biology of these cell types. 
\newline \newline
Now we can visualize a PC plot using only loops that are significantly
different between celltypes
```{r pca_sigfix, fig.width=7, fig.height=4, fig.cap = "PCA plot of differential loops"}
pcaPlot(topLoops(jnp_res, FDR = 0.05)) + geom_text(aes(label=samples)) 
```
Here, we see a much tighter cluster of the cell types, which may be expected
since the only variables introduced into the principal components are the loops
that are statistically different between the cell types. 

## Loop Plots
We can easily visualize the looping structure of a region specfied by a 
`GRanges` object in each sample. We also afford the annotation of the region
with the genes for a particular organism. Currently, mouse and human (default)
are supported.
```{r dloplots2,fig.width=7, fig.height=10}
regA <- GRanges(c("1"),ranges=IRanges(c(36000000),c(36300000)))
full.plot <- loopPlot(jn_res, regA)
```
$^1$Hnisz, Denes, et al. "Activation of proto-oncogenes by disruption of
chromosome neighborhoods." Science (2016): aad9024. <br><br>
$^2$Ji, Xiong, et al. "3D Chromosome Regulatory Landscape of Human Pluripotent
Cells." Cell stem cell (2015).
